---
title: "Lab 3"
author: "Tanner Huck"
date: "5/4/2022"
Collaborated with: "Xiangyi Wen"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1. t-test (10 points)
 Function: t-test
 Input: x, a numeric vector
        alternative, character string of alternative hypothesis
        mu, a number indicating null hypothesis value of mean
 Outputs: test_stat, numeric test statistic
          df, degrees of freedom
          alternative, value of parameter alternative
          p_val, the numeric p-value
```{r t-test function}
my_t.test <- function(x, alternative, mu) {
  # Check if parameter alternative is of the right form
  if (!(alternative == "two.sided" || "less" || "greater")) {
    stop("alternative must be of the form two.sided, less, or greater")
  }
  # Calculating the test statistic
  testStatistic <- (mean(x)-mu)/(sd(x)/sqrt(length(x)))
  # Calculating the degrees of freedom
  df <- length(x) - 1
  # Calculating the p value
  p_val <- 0
  if (alternative == "less") {
    p_val <- pt(testStatistic, df, lower.tail = TRUE, log.p = FALSE)
  } else if (alternative == "greater") {
    p_val <- pt(testStatistic, df, lower.tail = FALSE, log.p = FALSE)
  } else { 
    p_val <- 2 * pt(abs(testStatistic), df, lower.tail = FALSE, log.p = FALSE)
  }
  # Creating a list of the things I want to return
  my_list <- list(testStatistic,
                  df,
                  alternative,
                  p_val)
  # Returning values
  return(my_list)
}
```


```{r reading in helium data}
# reading in the helium data set
helium_data <- read.csv("https://www.openintro.org/data/csv/helium.csv")
```

```{r testing the t-test function}
# Comparing the t.test we made a function for compared to the real t.test
my_t.test(helium_data$helium, "two.sided", 20)
t.test(helium_data$helium, alternative = "two.sided", mu = 20)
```

As we can see when we compare the my_t.test function results to a t.test of the same data set, we get about the same answers. 

## Part 2. Linear model (10 points)
 Function: linear model
 Input: formula, a class object similar to lm
        data, data frame
 Outputs: a table similar to the coefficient table from summary
```{r linear fit function}
library(knitr)
library(kableExtra)
my_lm <- function(formula, data) {
  # Finding the coefficients of the linear regression
  x_matrix <- model.matrix(formula, data)
  model_frame <- model.frame(formula, data)
  y_matrix <- model.response(model_frame)
  lin_reg_coe <- solve(t(x_matrix)%*%(x_matrix))%*%t(x_matrix)%*%y_matrix
  # Finding the degrees of freedom
  df <- nrow(data) - 2
  # Estimating sigma squared
  sigma_squared <- sum((y_matrix-(x_matrix%*%lin_reg_coe))^2/df)
  # Estimating the standard error
  standard_error <- sqrt(diag(sigma_squared*solve(t(x_matrix)%*%(x_matrix))))
  # Calculating the t value
  t_val <- lin_reg_coe/standard_error
  # Calculating the p alues
  p_val_grade <- 2 * pt(abs(t_val), df, lower.tail = FALSE, log.p = FALSE)
  # Making the table that the function will return
  table_values <- cbind(lin_reg_coe, standard_error, t_val, p_val_grade)
  table <- kable(table_values, col.names = c("Estimate", "Std. Error", "t-value", "p-value"))
  return(table)
}
```

```{r reading in Duke data}
# reading in the Duke data set
grades_data <- read.csv("https://www.openintro.org/data/csv/gpa.csv")
```

```{r testing the linear model}
true_lm <- lm(gpa ~ studyweek, data = grades_data)
summary(true_lm)
my_lm(gpa ~ studyweek, grades_data)
```

As we can see when we compare the my_lm() function results to lm() of the same data set, we get the same coefficients (besides one of the p values being zero because of a very very small value). 





