---
title: "Short Lab 3"
author: "Tanner Huck"
date: "4-27-2022"
output: html_document
---

<!--- Begin styling code. --->
<style type="text/css">
/* Whole document: */
body{
  font-family: "Palatino Linotype", "Book Antiqua", Palatino, serif;
  font-size: 12pt;
}
h1.title {
  font-size: 38px;
  text-align: center;
}
h4.author {
  font-size: 18px;
  text-align: center;
}
h4.date {
  font-size: 18px;
  text-align: center;
}
</style>
<!--- End styling code. --->


As usual, all code below should follow the style guidelines from the lecture slides.

In this short lab, you will analyze data from the [Environmental Performance Index](https://epi.yale.edu/), which ranks 180 countries on 32 variables that describe environmental health and ecosystem vitality. We'll consider a subset of those indicators. See [this technical appendix](https://epi.yale.edu/downloads/epi2020technicalappendix20200803.pdf) (Page 2) for descriptions of the variables. 

The following code will load in the dataset and prepare it to run PCA on. 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
# load in environmental variables to run PCA on 
pca_df <- read_csv("https://raw.githubusercontent.com/svteichman/STAT302-WIN2022/main/files/projects/02_dimension_reduction/pca_data.csv")
pca_indicators <- pca_df %>% select(-country)
rownames(pca_indicators) <- pca_df$country
# load in additional data about countries
other_df <- read_csv("https://raw.githubusercontent.com/svteichman/STAT302-WIN2022/main/files/projects/02_dimension_reduction/additional_data.csv")
```

## Run PCA 

**1**. Run PCA on `pca_indicators` using the function `princomp()`. Don't worry about scaling the variables because all of the indicators are already on a 0-100 scale. 

```{r using PCA}
# using the princomp() function with the pca_indicators data
PCA <- princomp(pca_indicators)
```

**2**. Make a screeplot of your results from `princomp()` with the function `screeplot()`. What is an appropriate number of principal components to analyze? Why? 

```{r screenplot}
screeplot(PCA)
```

# I think we should analyze 2 principle components. This is because when you get past 2 components, the data is less important, or adding very little data. 

**3**. Plot your first two principal components (see slides 34 and 35 of Slides 5). Label between $5$ and $10$ countries. Interpret your results. Which countries have similar environmental indicators? Which countries have different environmental indicators? 

```{r}
plot_df <- 
  data.frame(pc1 = PCA$
               scores[, 1],
             pc2 = PCA$
               scores[, 2],
             name = rownames(pca_indicators))
ggplot(plot_df, 
       aes(x = pc1, y = pc2, 
           label = name)) + 
  geom_point() + 
  geom_text(size = 4, 
            nudge_x = .12, 
            nudge_y = .12) + 
  labs(x = "Principal Component 1", 
       y = "Principal Component 2",
       title = "First Two Principal Components of mtcars") + 
  theme_bw(base_size = 12) + 
  theme(plot.title = 
          element_text(hjust = 0.5))

text_df <- plot_df %>%
  filter(name %in%
           c("Iceland", "Canada",
             "Greenland", "United States of America",
             "Australia", "Viet Nam", "Russia"))
ggplot(plot_df, 
       aes(x = pc1, y = pc2,
           label = name)) + 
  geom_point() + 
  geom_text(data = text_df,
            size = 4, 
            nudge_x = .12, 
            nudge_y = .12) + 
  labs(x = "Principal Component 1", 
       y = "Principal Component 2",
       title = "First Two Principal Components of mtcars") + 
  theme_bw(base_size = 12) + 
  theme(plot.title = 
          element_text(hjust = 0.5))
```

# Out of the country that I chose to label, it looks like Canada, Iceland, and the US all have similar environmental indicators. Whereas and example of different environmental indicators would be Viet Nam and Australia.

**4**. Take you plot from (**3**) and color the points by one of the variables in `other_df`. Choose either `region`, `ldc` (which is $1$ for a set of least developed countries and $0$ otherwise), `emmrk` (which is $1$ if the country has an emerging market and $0$ otherwise), or `gdp` (if you choose `gpd` I recommend instead using `log(gdp)` because it is a very right-skewed variable). See slides 39 through 44 of Slides 5 for code inspiration. 

```{r}
ggplot(plot_df, 
       aes(x = pc1, y = pc2,
           label = name,
           color = as.factor(other_df$region))) +
  geom_point() + 
  labs(x = "Principal Component 1", 
       y = "Principal Component 2",
       title = "First Two Principal Components of mtcars") + 
  theme_bw(base_size = 12) + 
  theme(plot.title = 
          element_text(hjust = 0.5))
```

**5**. Describe the relationship between the first principal component and the variable `EPI_cat` in `other_df` (look at numerical and graphical measures of the relationship). `EPI_cat` is a weighted summary of all of the environmental indicators created by the people who built the Environmental Performance Index. Does this relationship make sense to you? Why or why not? 

The first pricncipal componenet is like a summary of the data, and excluding some of the less important data. The `EPI_cat` is doing something similar because it is a weighted summary of the enviromental indicators. I believe that these are similar becasuse they are both trying to summarize the enviromental indicators, and excluding some data is a unique way. However some of the principal component data is negative and all of the `EPI_cat` data is positive, which confuses me.

